---
title: "Baseline characteristics"
author: "[Sander W. van der Laan, PhD](https://vanderlaan.science) | s.w.vanderlaan@gmail.com"
date: "`r Sys.Date()`"
output:
  html_notebook:
    cache: yes
    code_folding: hide
    collapse: yes
    df_print: paged
    fig.align: center
    fig_caption: yes
    fig_height: 6
    fig_retina: 2
    fig_width: 7
    highlight: tango
    theme: lumen
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
mainfont: Arial
subtitle: Accompanying 'EndoMT_in_AE'
editor_options:
  chunk_output_type: inline
# bibliography: references.bib
# knit: worcs::cite_all
---

# General Setup
We will clean the environment, setup the locations, define colors, and create a datestamp.

_Clean the environment._
```{r echo = FALSE}
rm(list = ls())
```

_Set locations and working directories..._
```{r LocalSystem, echo = FALSE}
source("scripts/local.system.R")

```

_... a package-installation function ..._
```{r Function: installations}
source("scripts/functions.R")

```


_... and load those packages._
```{r loading_packages, message=FALSE, warning=FALSE}
source("scripts/pack01.packages.R")


```

_We will create a datestamp and define the Utrecht Science Park Colour Scheme_.
```{r Setting: Colors}

Today = format(as.Date(as.POSIXlt(Sys.time())), "%Y%m%d")
Today.Report = format(as.Date(as.POSIXlt(Sys.time())), "%A, %B %d, %Y")

source("scripts/colors.R")

```

```{r global_options, include = FALSE}
# further define some knitr-options.
knitr::opts_chunk$set(fig.width = 12, fig.height = 8, fig.path = 'Figures/', 
                      wwarning = TRUE, # show warnings during codebook generation
  message = TRUE, # show messages during codebook generation
  error = TRUE, # do not interrupt codebook generation in case of errors,
                # usually better for debugging
  echo = TRUE,  # show R code
                      eval = TRUE)
ggplot2::theme_set(ggplot2::theme_minimal())
pander::panderOptions("table.split.table", Inf)
```

# Introduction

Here we load in the target list, MAGMA results from GWAS derived gene-based association studies, and lookup the target genes. 

# Load targets

```{r}
library(openxlsx)
endo_targets <- openxlsx::read.xlsx("clusters/20230525_Midstage_genes_10323_EndMTpaper.xlsx", sheet = "Targets")
target_list <- unlist(endo_targets$GeneID)

```

# Load MAGMA results

# CAD

```{r}

DATASETS <- list("CAC", "CHARGE_CAC_EA_AA", "cIMT", "Plaque", "CAD", "CES", "IS", "LAS", "AS", "SVD", "AF", "FrmrSmk", "EvrSmk", "CpD", "logOnset", "T2D", "T2DadjBMI", "BMI", "TC", "TG", "LDL", "HDL", "SBP", "DBP", "PP", "MDD", "SWB", "BIP", "Insomnia", "Neuroticism")

# DATASETS <- list("CAC")

magma.results <- data.frame(matrix(NA, ncol = 10, nrow = 0))

magma.import <- function(tempDF, trait, target, verbose=c(TRUE,FALSE)){
  cat("> collecting data...\n\n")
    Symbol = tempDF$SYMBOL
    EnsemblID = tempDF$GENE
    Chr = tempDF$CHR
    Start = tempDF$START
    Stop = tempDF$STOP
    Nsnps = tempDF$NSNPS
    Nsamples = tempDF$N
    Zstat = tempDF$ZSTAT
    Pvalue = tempDF$P
    
    output = c(trait, EnsemblID, Symbol, Chr, Start, Stop, Nsnps, Nsamples, Zstat, Pvalue)
    
    if (verbose == TRUE) {
      cat("We have collected the following and summarize it in an object:\n")
      cat("Phenotype.....:", trait, "\n")
      cat("Symbol........:", Symbol, "\n")
      cat("EnsemblID.....:", EnsemblID, "\n")
      cat("Chr...........:", Chr, "\n")
      cat("Start.........:", Start, "\n")
      cat("Stop..........:", Stop, "\n")
      cat("Nsnps.........:", Nsnps, "\n")
      cat("Nsamples......:", Nsamples, "\n")
      cat("Zstat.........:", round(Zstat, 6), "\n")
      cat("P.............:", signif(Pvalue, 8), "\n")
    } else {
      cat("Collecting data in summary object.\n")
    }
  return(output)
  print(output)
}


for (target_of_interest in 1:length(target_list)) {
  TARGET = target_list[target_of_interest]
  for (TRAIT in DATASETS) {
    cat(paste0("Getting data for ["), TARGET, "] in ",TRAIT,".\n")
    currentDF <- fread(paste0(GWAS_loc, "/_magma/",TRAIT,"/",TRAIT,".magma.genes.out")) %>%
      dplyr::filter(., SYMBOL==TARGET)
    # for debug
    # print(DT::datatable(currentDF))
    print(nrow(currentDF))
    # print(str(currentDF))
    
    magma.results.TEMP <- data.frame(matrix(NA, ncol = 10, nrow = 0))
    magma.results.TEMP[1,] = magma.import(currentDF, TRAIT, TARGET, verbose = F)
    magma.results = rbind(magma.results, magma.results.TEMP)
  
  }
}

cat("Edit the column names...\n")
colnames(magma.results) = c("Phenotype", "EnsemblID", "Symbol",
                            "Chr", "Start", "Stop",
                            "Nsnps", "Nsamples",
                            "Zstat", "Pvalue")
cat("Correct the variable types...\n")
magma.results$Phenotype <- as.character(magma.results$Phenotype)
magma.results$EnsemblID <- as.character(magma.results$EnsemblID)
magma.results$Symbol <- as.character(magma.results$Symbol)
magma.results$Chr <- as.numeric(magma.results$Chr)
magma.results$Start <- as.numeric(magma.results$Start)
magma.results$Stop <- as.numeric(magma.results$Stop)
magma.results$Nsnps <- as.numeric(magma.results$Nsnps)
magma.results$Nsamples <- as.numeric(magma.results$Nsamples)
magma.results$Zstat <- as.numeric(magma.results$Zstat)
magma.results$Pvalue <- as.numeric(magma.results$Pvalue)



# Removing intermediates
cat("Removing intermediate files...\n")
rm(currentDF, TRAIT, TARGET, magma.results.TEMP)

```

```{r}
magma.results.filt <- magma.results %>% dplyr::filter(., !is.na(Pvalue))

# Save the data
cat("Writing results to Excel-file...\n")
library(openxlsx)
write.xlsx(magma.results.filt,
           file = paste0(OUT_loc, "/",Today,".EndoMT.targets.GWASmagma.xlsx"),
           rowNames = FALSE, colNames = TRUE, sheetName = "MAGMA_results")

```

# Visualize
```{r}
# install.packages.auto("CMplot")
library("CMplot")
data(pig60K)
head(pig60K)

setwd(paste0(PLOT_loc))
for (TRAIT in DATASETS) {
  temp <- subset(magma.results.filt, 
                 Phenotype == TRAIT, 
                 select = c("Symbol", "Chr", "Start", "Pvalue"))
  highlight_list <- as.vector(subset(temp, Pvalue <= 1e-6, select = c("Symbol")))
  # str(highlight_list)
  CMplot(temp, plot.type = "m", col = c("grey30", "grey60"), LOG10 = TRUE, 
         ylim = c(0,50),
         threshold = c(1e-6, 0.05), threshold.lty = c(1,2), threshold.lwd = c(1,1), threshold.col = c("black","grey"), amplify=TRUE,
         chr.den.col = NULL, 
         highlight = highlight_list, highlight.text = highlight_list,
         signal.col = c(uithof_color[3], uithof_color[16]), signal.cex = c(1.5,1.5), signal.pch = c(19,19),
         main = TRAIT,
         file = "pdf", file.name = paste0(Today,".",TRAIT), dpi = 300, file.output = TRUE, 
         verbose = FALSE, 
         width = 14, height = 6)
}
setwd(paste0(PROJECT_loc))
rm(temp, highlight_list)
```


# Session information

------------------------------------------------------------------------------------------------------------------------

    Version:      v1.0.1
    Last update:  2023-05-26
    Written by:   Sander W. van der Laan (s.w.vanderlaan-2[at]umcutrecht.nl).
    Description:  Script to get some Athero-Express Biobank Study baseline characteristics.
    Minimum requirements: R version 3.4.3 (2017-06-30) -- 'Single Candle', Mac OS X El Capitan

    **MoSCoW To-Do List**
    The things we Must, Should, Could, and Would have given the time we have.
    _M_

    _S_

    _C_

    _W_

    **Changes log**
    * v1.0.1 Got MAGMA results. Added to Excel-file. Visualizations. 
    * v1.0.0 Initial version. 

------------------------------------------------------------------------------------------------------------------------

```{r eval = TRUE}
sessionInfo()
```

# Saving environment

```{r Saving}
save.image(paste0(PROJECT_loc, "/",Today,".",PROJECTNAME,".EndoMT_in_AE.RData"))
```

+-----------------------------------------------------------------------------------------------------------------------------------------+
| <sup>© 1979-2023 Sander W. van der Laan | s.w.vanderlaan[at]gmail.com | [vanderlaan.science](https://vanderlaan.science).</sup> |
+-----------------------------------------------------------------------------------------------------------------------------------------+
